Assessing the Replicability of Psychological Science Through Pedagogy

### 2016-09-08 paper/paper.Rmd
Pasting error
minor updates

author-information:
    - \author{Eric N. Smith*, Robert X. D. Hawkins*, the students of Psych 254, and Michael C. Frank}

@@

Replicability is a core value for empirical research and there is increasing concern throughout psychology that more independent replication is necessary [@osc2015;@wagenmakers2012]. Yet under the current incentive structure for science, direct replication is not typically valued. One potential solution to this problem is to make direct replication an explicit part of pedagogy: that is, to teach students about experimental methods by asking them to run replication studies [@frank2012;@grahe2012]. Despite enthusiasm for this idea [@lebel2015;@everett2015;@standing2016;@king2016], there is limited data beyond anecdotal reports or individual projects [@phillips2015;@lakens2013] to support its efficacy.

@@

It is a challenging proposition to estimate the replicability of an entire literature [cf. @osc2015]. Making such an estimate minimally requires determining a sampling strategy, a notion of what constitutes a direct replication, and a standard for what constitutes a successful replication, among many other difficult decisions [@gilbert2016;@anderson2016]. In the current work, we focus on a quantity that is easier to estimate: the probability that a graduate student can choose an article of interest in *Psychological Science* and -- in a single attempt, within constraints of budget, expertise, and effort -- reproduce the basic pattern of findings with sufficient fidelity to be able to build on it in future work. We believe that this notion of whether a piece of research supports cumulative progress is an important construct. Whether an independent scientist can build on a published result is perhaps as important as whether the work is replicable in a more conceptual sense -- and in many cases easier to determine.

With respect to pedagogy, we describe our process for conducting well-powered, high-quality, direct replications as part of classroom pedagogy so that other institutions may adapt these practices. This process is dramatically facilitated by the use of free standardized tools for statistical analysis (R), data and code sharing (Git, R Markdown), and creation of web experiments (HTML, CSS, JavaScript). Nevertheless, there are significant limitations on what can be done in a single term and within the constraints of a course budget. We return in the Discussion to limitations on our approach.

In sum, we view the contribution of the current work to be two-fold. First, we provide a proof-of-concept and pedagogical model for conducting replications in the classroom. Second, we provide an initial estimate of what proportion of studies from the most recent volume of *Psychological Science* could be reproduced -- supporting the same theoretical claims -- by a motivated student.

@@

## Study Selection

All projects were completed as part of a graduate-level methodology class (syllabus and materials available at [http://psych254.stanford.edu]()). At the initiation of class, all students were told that they had the opportunity to contribute their individual class assignment to a group replication project. The requirements for joining the project were to conduct a pre-registered replication of a finding from the 2015 volume of *Psychological Science* and to contribute the code, data, and materials to the writeup.

Students selected projects based on interest, rather than via a systematic sampling strategy. After selecting a paper of interest, students chose a particular study from within that paper based on their judgment of interest and feasibility. The modal study chosen was the first, but several students chose later studies in a set as well.  Thus, all interpretation of our findings should be with respect to the distribution of student interest. Any reproducibility estimates arising from our analyses are estimates of the probability of reproducing a finding that is of interest, rather than findings in general. This sampling strategy may thus bias our results towards more exciting or novel findings, as well as those that were more feasible for students. A summary of the original experiments chosen for replication is given in Table \ref{tab:summary}.

@@

Determining appropriate sample sizes is a major challenge in replication research [@button2013;@simonsohn2015]. Post-hoc power analyses are problematic because of the likelihood of inflation of effect sizes due to the "winner's curse" [@hoenig2001;@button2013], but -- especially within the constraints of a limited budget -- it can be impossible to follow more conservative guidelines in all cases. For example, @simonsohn2015 recommends 2.5x the original sample, which can be feasible for small or under-powered original studies but may lead to impractical or unnecessary recommendations for experiments that were initially large and/or adequately powered.

We eventually used a hybrid, case-by-case selection criterion that attempted to maximize the success of the project while staying within the constraints of our budget.
^[We initially were allocated \$1,500 but one student contributed personal research funds, leading to a total cost of $\approx$ \$1,700 for all studies. We attempted to set payment for our studies at approximately \$6/hour based on timing estimated from Pilot A sessions.] 
In 
`r sum(project_plan$power_standard == "80% power")`
cases, we powered the replication attempt to 80\% post-hoc power. In 
`r sum(project_plan$power_standard == "Original")` 
cases, we used the original sample size. Three replications deviated from these two criteria.  In one case, the original experiment was extremely over-powered in the post-hoc analysis, so we chose a smaller sample size that still would yield high power.  In another, we reduced the number of trials as part of the adaptation to Mechanical Turk. In the last case of a multi-step procedure where the first step consisted of stimulus generation, we decreased the N in the first stage of the procedure (details available in SOM).

@@

Students wrote their final reports using [R Markdown](http://rmarkdown.rstudio.com/), a "literate programming" environment in which statistical analysis code can be interspersed with text, figures, and tables. This part of our workflow was extremely useful both pedagogically and for encouraging reproducible research practices. Using the template developed by @osc2015 for replication reports, students created dynamic documents that included their proposed replication study, the code necessary to analyze their data, and the outputs of their data analysis (e.g. statistical tests, figures). This method of writing allows students to share a single compiled document via a hyperlink, facilitating review of writing, results, and code together in a single platform.^[The current manuscript is written in this fashion as well.] This writing method is also likely to reduce the frequency of statistical reporting errors [which appear to be regrettably common; @nuijten2015], given that many of these are likely introduced by transferring results between analysis outputs and manuscripts.

@@

In addition to as-needed guidance on projects, to ensure high quality replications, students went through a rigorous process in which each student's work was reviewed several times by both the instructor (MCF) and the teaching assistants (ENS and RXDH). Each student collected two pilot samples and review was coordinated with these. The first, "Pilot A," consisted of a handful of non-naive participants (e.g., the experimenter, other students). The goal of this pilot was to ensure that all needed data were being logged by the experiment script and that analytic scripts for the confirmatory analyses functioned appropriately. After Pilot A was completed, the instructor or a TA critiqued and reviewed the student's experimental script (acting as a participant), analytic code, and resulting output.

@@

## Statistical approach

@@

@@

Of the 15 students in the class, all completed projects (two with extensions going beyond the class period), one chose a project outside of the sampling frame, and three opted out of the broader project prior to data collection.^[Note that these students did not express doubts about the success of their projects prior to opting out, and 2/3 of these projects were judged by the instructor team to have been successful replications. Thus we have no evidence that these students opted out systematically due to their belief that their project would fail.] The remaining 11 students contributed code, data, registrations and materials to the final product. (In one case, a student did not complete the pre-registration procedure correctly but nevertheless submitted a pre-specified analysis plan to the instructors for review.  Thus, this project remains included.)

Each member of the instructor team coded the fidelity of replications, weighing whether the materials, sample, and task parameters differed from the original studies. Coded on a scale of 1 -- 7 (with 1 being a loose replication with substantive deviations from the original, and 7 being essentially identical), the mean rating was `r round(mean(d$fidelity), digits = 2)` (Range = `r round(min(d$fidelity), digits = 2)` to `r round(max(d$fidelity), digits = 2)`). Several studies were essentially identical to the originals, but there was a group of others that included differences in population or number of trials that might have an effect on the results.

@@

Students of the class and the instructors gave independent “replication ratings” to assess subjective judgement of the replication. Our three-point rating system was based on the theoretical interpretation of the original finding, with 0\% = no support for the original interpretation, 50\% = some support for the original interpretation, 100\% = full replication). We found that the ratings of the student carrying out the replication and the instructors were in agreement for `r sum(d$replicated_TA == d$replicated_student,na.rm=T)` out of the `r nrow(d)` replications.  The average rating given by instructors and students was `r round(mean(d$replicated_TA, na.rm=TRUE), digits = 2)` and `r round(mean(d$replicated_student, na.rm=TRUE), digits = 2)` respectively.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rl|ccccccc}
  \hline
\multicolumn{2}{c}{Support} & \multicolumn{2}{c}{Original on Turk?} & \multicolumn{2}{c}{High fidelity?} & \multicolumn{3}{c}{Open Materials?} \\
  \hline
   &  & No & Yes & Yes & No & No & Some & Yes \\ 
  Full & 3 & 1 & 2 & 1 & 2 & 1 & 1 & 1 \\ 
  Partial & 6 & 4 & 2 & 4 & 2 & 2 & 1 & 3 \\ 
  No &  2 &  1 &  1 &  2 & 0 &  1 & 0 &  1 \\ 
  \hline
  Total & 11 & 6 & 5 & 7 & 4 & 4 & 2 & 5 \\
   \hline
\end{tabular}
\caption{\label{tab:results} Counts of instructor team ratings of replication support for the theoretical findings, split by variables of interest. High fidelity denotes replications judged to be above the median rating of fidelity ( >`r median(d$fidelity)` out of 7).}
\end{table}

@@

These additional measures therefore give another sense in which replications can be successful: if we find roughly the same effect size or Bayes factor despite $p$-values on different sides of the $p = 0.05$ threshold, we should have increased confidence in the effect's reproducibility. To compute a standardized effect size, we followed @osc2015 in converting test statistics to a measure of "correlation coefficient per df," which is bounded between 0 and 1^[For example, the conversion from a $t$ statistic is given by $r = \sqrt{t^2/(t^2 + df)}$, where $df$ is the degrees of freedom. Other formulas are given in @osc2015, Supplemental Information.]. We first compared effect sizes from the original and replication studies (Figure 2A). Effect sizes were highly correlated between the two sets of studies 
(r = `r signif(cor.test(d$original_ES_d,d$rep_es)$estimate, digits =2)`, p < .0001), but they were generally smaller in the replications 
(`r signif(coef(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100`\% of the original; 95\% CI [`r signif(confint(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100` - `r signif(confint(lm(d$rep_es~d$original_ES_d-1))[2], digits = 2)*100`]).

While we preregistered an analysis determining whether the original reported effect size is included in the 95\% confidence interval found in the replication, we instead opted for a straightforward comparison of point estimates. This choice was motivated by recent concerns that using confidence intervals in this manner may lead to misleading interpretations [@morey2016], as well as discrepencies in the authors' reporting of effect sizes across the articles we included (which made computing confidence intervals difficult in some cases).

@@

p1 <- ggplot(d,aes(x=original_ES_d,y=rep_es)) +
  geom_point(alpha = .8, 
             aes(color=replicated_TA_factor, size = collected_n)) +
  geom_text_repel(aes(label = first_author_name), size = 2, force = 2) +
  scale_x_continuous(limits = c(0, 1), 
                breaks = c(.1, .3, .5, .7, .9)) +
  scale_y_continuous(limits = c(0, 1), 
                breaks = c(.1, .3, .5, .7, .9)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_smooth(method = "lm", se=FALSE) +
  xlab("Orig. effect size (d)") + 
  ylab("Rep. effect size (d)") + 
  theme_bw(base_size=10) + theme(legend.position="bottom") + 
  scale_size(name = "Rep. N") +
  # scale_color_solarized(guide = FALSE)
  scale_color_solarized(name = "Subjective rep. rating")

@@

<!-- Demographics -->

@@

In this paper, we reported the results of a series of 11 direct replications of previously-published experiments from the 2015 volume of *Psychological Science*. We had two interlocking goals: first, to create a proof of concept for high-quality pedagogical replications, and second, to provide evidence on the ability of students replicate important, recent findings in a top journal. Despite our relatively tight adherence to the original experimental protocols and relatively large sample sizes, our results were underwhelming and further suggests the need for more replication attempts as a means for cumulative scientific progress. The modal outcome of our projects was partial replication, in which some hint of the original pattern was observed but the key statistical test was not statistically significant and showed both smaller effect size and lower evidential value than the same test in the original.

Our project aimed to assess the probability that a finding -- selected on grounds of feasibility and interest -- would be replicated with the time, money, and effort constraints of a course project. While these constraints might initially seem severe, our work here shows what is possible for motivated students using freely-available tools and resources. Indeed, the majority of the projects in our report were relatively close replications and included the same or greater numbers of participants. Nevertheless, they had a number of limitations, including at least: 1) limited time for iterative piloting and adjustment [@lewis2016], 2) limits on funding for samples, and 3) limits on domain expertise with respect to the specific effects that were chosen. In addition, many papers published within our notional sampling frame (the 2015 volume of *Psychological Science*) were not feasible as projects, and students in our course were asked to take feasibility into account in choosing their projects as well as their own interests.

>>>>

author-information:
    - \author{Eric Smith*, Robert X. D. Hawkins*, the students of Psych 254, and Michael C. Frank}

@@

Replicability is a core value for empirical research and there is increasing concern throughout psychology that more independent replication is necessary [@osc2015;@wagenmakers2012]. Under the current incentive structure for science, direct replication is not typically valued, however. One potential solution to this problem is to make direct replication an explicit part of pedagogy: that is, to teach students about experimental methods by asking them to run replication studies [@frank2012;@grahe2012]. Despite enthusiasm for this idea, however [@lebel2015;@everett2015;@standing2016;@king2016], there is limited data -- beyond anecdotal reports or individual projects [@phillips2015;@lakens2013] -- to support its efficacy.

@@

It is a challenging proposition to estimate the replicability of an entire literature [cf. @osc2015]. Making such an estimate minimally requires determining a sampling strategy, a notion of what constitutes a direct replication, and a standard for what constitutes a successful replication, among many other difficult decisions [@gilbert2016;@anderson2016]. In the current work, we focus on a quantity that is easier to estimate: the probability that a graduate student can choose an article of interest in *Psychological Science* and -- in a single attempt, within constraints of budget, expertise, and effort -- reproduce the basic pattern of findings with sufficient fidelity to be able to build on it in future work. We believe that this notion of whether a piece of research supports cumulative progress is an important construct. Whether another independent scientist can build on a result is perhaps as important as -- and in many cases easier to measure -- whether the work is replicable in some more abstract sense.

With respect to pedagogy, we describe our process for conducting well-powered, high-quality, direct replications as part of classroom pedagogy. This process is dramatically facilitated by the use of standardized (and free) technical tools for statistical analysis, data and code sharing, and creation of web experiments. Nevertheless, there are significant limitations on what can be done in a single term and within the constraints of a course budget. We return in the Discussion to limitations on our approach.

In sum, we view the contribution of the current work to be two-fold. First, we provide a proof-of-concept and pedagogical model for conducting replications in the classroom. Second, we provide an initial estimate of what proportion of studies from the most recent volume of *Psychological Science* could be reproduced -- at least as far as support for the same theoretical claims -- by a motivated student.

@@

## Class Outline

All projects were completed as part of a class (syllabus and materials available at [http://psych254.stanford.edu]()). At the initiation of class, all students were told that they had the opportunity to participate in a group replication project to which they could contribute their class project. The requirements for joining the project were to conduct a pre-registered replication of a finding from the 2015 volume of *Psychological Science* and to contribute the code, data, and materials to the writeup.

Students selected projects based on interest, rather than via a systematic sampling strategy. Thus, all interpretation of our findings should be with respect to the distribution of student interest. Any reproducibility estimates arising from our analyses are estimates of the probability of reproducing a finding that is of interest, rather than findings in general. This sampling strategy may thus bias our results towards more exciting or novel findings, as well as those that were more feasible for students. A summary of the original findings is given in Table \ref{tab:summary}.

@@

@@

Students wrote their final reports using [R Markdown](http://rmarkdown.rstudio.com/), a "literate programming" environment in which statistical analysis code can be interspersed with text, figures, and tables. This part of our workflow was extremely useful both pedagogically and for encouraging reproducible research practices. Using the template developed by @osc2015 for replication reports, students created dynamic documents that included their proposed replication study, the code necessary to analyze their data, and the outputs of their data analysis (e.g. statistical tests, figures). These documents could be compiled into a variety of formats (e.g., HTML, PDF, Microsoft Word). This method of writing allows students to share a single compiled document via a hyperlink, facilitating review of writing, results, and code together in a single platform.^[The current manuscript is written in this fashion as well.] This writing method is also likely to reduce the frequency of statistical reporting errors [which appear to be regrettably common; @nuijten2015], given that many of these are likely introduced by copying and pasting between analysis outputs and manuscripts.

@@

In addition to as-needed guidance on projects, to ensure high quality replications, students went through a rigorous process in which each student's work was reviewed several times by both the instructor (MCF) and the teaching assistants (ES and RXDH). Each student collected two pilot samples and review was coordinated with these. The first, "Pilot A," consisted of a handful of non-naive participants (e.g., the experimenter, other students). The goal of this pilot was to ensure that all needed data were being logged by the experiment script and that analytic scripts for the confirmatory analyses functioned appropriately. After Pilot A was completed, the instructor or a TA reviewed the student's experimental script (acting as a participant) and critiqued and reviewed the student's full analytic code and output.

@@

## Sample size determination and statistical approach

After selecting a paper of interest, students chose a particular study from within that paper based on their judgment of interest and feasibility. The modal study chosen was the first, but several students chose later studies in a set as well.

@@

Determining appropriate sample sizes is a second major challenge in replication research [@button2013;@simonsohn2015]. Post-hoc power analyses are problematic because of the likelihood of inflation of effect sizes due to the "winner's curse" [@hoenig2001;@button2013], but -- especially within the constraints of a limited budget -- it can be impossible to follow more conservative guidelines in all cases. For example, @simonsohn2015 recommends 2.5x the original sample, which can be feasible for small or under-powered original studies but may lead to impractical or unnecessary recommendations for experiments that were initially large and/or adequately powered.

We eventually used a hybrid, case-by-case selection criterion that attempted to maximize the success of the project while staying within the constraints of our budget.
^[We initially were allocated \$1,500 but one student contributed personal research funds, leading to a total cost of $\approx$ \$1,700 for all studies. We attempted to set payment for our studies at approximately \$6/hour based on timing estimated from Pilot A sessions.] 
In 
`r sum(project_plan$power_standard == "80% power")`
cases, we powered the replication attempt to 80\% post-hoc power. In 
`r sum(project_plan$power_standard == "Original")` 
cases, we used the original sample size. In one case, the original experiment was  extremely over-powered in the post-hoc analysis, so we chose a smaller sample size that still would yield high power.  In one case, we reduced the number of trials as part of the adaptation to Mechanical Turk, and in one case of a multi-step procedure where the first step consisted of stimulus generation, we decreased the N in the first stage of the procedure (details available in SOM).

## Individual Replications

@@

Of the 15 students in the class, all completed projects (two with extensions, one short and one more significant), but one failed to find a project of interest in the sampling frame and three opted out of the broader project prior to data collection.^[Note that these students did not express doubts about the success of their projects prior to opting out, and 2/3 of these projects were judged by the instructor team to have been successful replications. Thus we have no evidence that these students opted out systematically due to their belief that their project would fail.] The remaining 11 students contributed code, data, registrations and materials to the final product. (In one case, a student failed to complete the pre-registration procedure correctly but nevertheless submitted a pre-specified analysis plan to the instructors for review).

Each member of the instructor team coded the fidelity of replications, weighing whether the materials, sample, and task parameters differed from the original studies. Coded on a scale of 1 -- 7 (with 1 being a loose replication with substantive deviations from the original, and 7 being essentially identical), the mean rating was `r round(mean(d$fidelity), digits = 2)`. Several studies were essentially identical to the originals, but there was a group of others that included differences in population or number of trials that might have an effect on the results.

@@

Students of the class and the instructors gave independent “replication ratings” to assess subjective judgement of the replication. Our rating system was based on the theoretical interpretation of the original finding, with 0 = no support for the original interpretation, .5 = some support for the original interpretation, 1 = full replication). We found that the ratings of the student carrying out the replication and the instructors were in agreement for `r sum(d$replicated_TA == d$replicated_student,na.rm=T)` out of the `r nrow(d)` replications.  The average rating given by instructors and students was `r round(mean(d$replicated_TA, na.rm=TRUE), digits = 2)` and `r round(mean(d$replicated_student, na.rm=TRUE), digits = 2)` respectively.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rl|ccccccc}
  \hline
\multicolumn{2}{c}{Replicated} & \multicolumn{2}{c}{Original on Turk?} & \multicolumn{2}{c}{High fidelity?} & \multicolumn{3}{c}{Open Materials?} \\
  \hline
   &  & No & Yes & Yes & No & No & Some & Yes \\ 
  Full & 3 & 1 & 2 & 1 & 2 & 1 & 1 & 1 \\ 
  Partial & 6 & 4 & 2 & 4 & 2 & 2 & 1 & 3 \\ 
  No &  2 &  1 &  1 &  2 & 0 &  1 & 0 &  1 \\ 
  \hline
  Total & 11 & 6 & 5 & 7 & 4 & 4 & 2 & 5 \\
   \hline
\end{tabular}
\caption{\label{tab:results} Counts of instructor team replication ratings, split by mediating variables.}
\end{table}

@@

These additional measures therefore give another sense in which replications can be successful: if we find roughly the same effect size or Bayes factor despite $p$-values on different sides of the $p = 0.05$ threshold, we should have increased confidence in the effect's reproducibility. To compute a standardized effect size, we followed @osc2015 in converting test statistics to a measure of "correlation coefficient per df," which is bounded between 0 and 1^[For example, the conversion from a $t$ statistic is given by $r = \sqrt{t^2/(t^2 + df)}$, where $df$ is the degrees of freedom. Other formulas are given in @osc2015, Supplemental Information]. We first compared effect sizes from the original and replication studies (Figure 2A). Effect sizes were highly correlated between the two sets of studies 
(r = `r signif(cor.test(d$original_ES_d,d$rep_es)$estimate, digits =2)`, p < .0001), but they were generally smaller in the replications 
(`r signif(coef(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100`\% of the original [`r signif(confint(lm(d$rep_es~d$original_ES_d-1))[1], digits = 2)*100` - `r signif(confint(lm(d$rep_es~d$original_ES_d-1))[2], digits = 2)*100`]).

While we preregistered an analysis determining whether the original reported effect size is included in the 95\% confidence interval found in the replication, we instead opted for a straightforward comparison of point estimates. This choice was motivated by recent discussions of the utility and validity of interpreting confidence intervals in this manner [@morey2016], as well as discrepencies in the reporting of effect sizes across the articles we included (which made computing confidence intervals difficult in some cases).

@@

p1 <- ggplot(d,aes(x=original_ES_d,y=rep_es)) +
  geom_point(alpha = .8, 
             aes(color=replicated_TA_factor, size = collected_n)) +
  geom_text_repel(aes(label = first_author_name), size = 2, force = 2) +
  scale_x_continuous(limits = c(0, 1), 
                breaks = c(.1, .3, .5)) +
  scale_y_continuous(limits = c(0, 1), 
                breaks = c(.1, .3, .5)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_smooth(method = "lm", se=FALSE) +
  xlab("Orig. effect size (d)") + 
  ylab("Rep. effect size (d)") + 
  theme_bw(base_size=10) + theme(legend.position="bottom") + 
  scale_size(name = "Rep. N") +
  # scale_color_solarized(guide = FALSE)
  scale_color_solarized(name = "Subjective rep. rating")

@@

@@

In this paper, we reported the results of a series of 11 direct replications of previously-published experiments from the 2015 volume of *Psychological Science*. We had two interlocking goals: first, to create a proof of concept for high-quality pedagogical replications, and second, to provide evidence on the ability of students replicate important, recent findings in a top journal. Despite our relatively tight adherence to the original experimental protocols and relatively large sample sizes, our results were disappointing. The modal outcome of our projects was partial replication, in which some hint of the original pattern was observed but the key statistical test was not statistically significant and showed both smaller effect size and lower evidential value than the same test in the original.

Our project aimed to assess the probability that a finding -- selected on grounds of feasibility and interest -- would be replicated with the time, money, and effort constraints of a course project. While these constraints might initially seem severe, our work here shows what is possible for motivated students using freely-available tools and resources. Indeed, the majority of the projects in our report were relatively close replications and included the same or greater numbers of participants. Nevertheless, they had a number of limitations, including at least: 1) limited time for iterative piloting and adjustment [@lewis2016], 2) limits on available samples, and 3) limits on domain expertise with respect to the specific effects that were chosen. In addition, many papers published within our notional sampling frame (the 2015 volume of *Psychological Science*) were not feasible as projects, and students in our course were asked to take feasibility into account in choosing their projects as well as their own interests.
